---
title: "IV: Post-processing and passive imputation"
author: "Gerko Vink and Stef van Buuren"
date: "**Practical 4 of 4**"
output: html_document
---

This is the fourth vignette in the Winnipeg series. 

In this vignette we will walk you through the more advanced features of `mice`, such as *post-processing* of imputations and *passive imputation*.

---

**1. Open `R` and load the packages `mice` and `lattice`.**
```{r message=FALSE, warning=FALSE}
require(mice)
require(lattice)
set.seed(123)
```
We choose seed value `123`. This is an arbitrary value; any value would be an equally good seed value. Fixing the random seed enables you (and others) to exactly replicate anything that involves random number generators. If you set the seed in your `R` instance to `123`, you will get the exact same results and plots as we present in this document. 

---



**Post-proccessing of the imputations**

Remember that we imputed the `boys` data in the previous tutorial with `pmm` and with `norm`. One of the problems with the imputed values of `tv` with `norm` is that there are negative values among the imputations. Somehow we should be able to lay a constraint on the imputed values of `tv`. 

The `mice()` function has an argument called `post` that takes a vector of strings of `R` commands. These commands are parsed and evaluated after the univariate imputation function returns, and thus provides a way of post-processing the imputed values while using the processed version in the imputation algorithm. In other words; the post-processing allows us to manipulate the imputations for a particular variable that are generated within each iteration. Such manipulations directly affect the imputed values of that variable and the imputations for other variables. Naturally, such a procedure should be handled with care. 

---

**2. Post-process the values to constrain them between 1 and 25, use `norm` as the imputation method for `tv`.**
```{r}
ini <- mice(boys, maxit = 0)
meth <- ini$meth
meth["tv"] <- "norm"
post <- ini$post
post["tv"] <- "imp[[j]][, i] <- squeeze(imp[[j]][, i], c(1, 25))"
imp <- mice(boys, meth=meth, post=post, print=FALSE)
```
In this way the imputed values of `tv` are constrained (squeezed by function `squeeze()`) between 1 and 25.

---

**3. Compare the imputed values and histograms of `tv` for the solution obtained by `pmm` to the constrained solution (created with `norm`, constrained between 1 and 25).**

First, we recreate the default `pmm` solution
```{r}
imp.pmm <- mice(boys, print=FALSE)
```
and we inspect the imputed values for the `norm` solution
```{r}
table(complete(imp)$tv)
```
and for the `pmm` solution
```{r}
table(complete(imp.pmm)$tv)
```
It is clear that the norm solution does not give us integer data as imputations. Next, we inspect and compare the density of the incomplete and imputed data for the constrained solution.
```{r}
densityplot(imp, ~tv)
```

---

A nice way of plotting the histograms of both datasets simultaneously is by creating first the dataframe (here we named it `tvm`) that contains the data in one column and the imputation method in another column.
```{r}
tv <- c(complete(imp.pmm)$tv, complete(imp)$tv)
method <- rep(c("pmm", "norm"), each = nrow(boys))
tvm <- data.frame(tv = tv, method = method)
```
and then plotting a histogram of `tv` conditional on method.
```{r}
histogram( ~tv | method, data = tvm, nint = 25)
```

Is there still a difference in distribution between the two different imputation methods? Which imputations are more plausible do you think?

---

**4. Make a missing data indicator (name it `miss`) for `bmi` and check the relation of `bmi`, `wgt` and `hgt` for the boys in the imputed data. To do so, plot the imputed values against their respective calculated values.**
```{r}
miss <- is.na(imp$data$bmi)
xyplot(imp, bmi ~ I (wgt / (hgt / 100)^2),
       na.groups = miss, cex = c(0.8, 1.2), pch = c(1, 20),
       ylab = "BMI (kg/m2) Imputed", xlab = "BMI (kg/m2) Calculated")
```

With this plot we show that the relation between `hgt`, `wgt` and `bmi` is not preserved in the imputed values. In order to preserve this relation, we should use passive imputation.

---

**Passive Imputation**

There is often a need for transformed, combined or recoded versions of the data. In the case of incomplete data, one could impute the original, and transform the completed original afterwards, or transform the incomplete original and impute the transformed version. If, however, both the original and the transformed version are needed within the imputation algorithm, neither of these approaches work: One cannot be sure that the transformation holds between the imputed values of the original and transformed versions. `mice` has a built-in approach, called *passive imputation*, to deal with situations as described above. 

The goal of passive imputation is to maintain the consistency among different transformations of the same data. As an example, consider the following deterministic function in the `boys` data
\[\text{BMI} = \frac{\text{Weight (kg)}}{\text{Height}^2 \text{(m)}}\].


**5. Use passive imputation to conserve the relation between imputed `bmi`, `wgt` and `hgt` by setting the imputation method for `bmi` to `meth["bmi"]<- "~ I(wgt / (hgt / 100)^2)"`.**
```{r}
meth<- ini$meth
meth["bmi"]<- "~ I(wgt / (hgt / 100)^2)"
imp <- mice(boys, meth=meth, print=FALSE)
```

---

**6. Again, plot the imputed values of `bmi` versus the calculated values and check whether there is convergence for `bmi`.**

To inspect the relation:
```{r}
xyplot(imp, bmi ~ I(wgt / (hgt / 100)^2), na.groups = miss,
       cex = c(1, 1), pch = c(1, 20),
       ylab = "BMI (kg/m2) Imputed", xlab = "BMI (kg/m2) Calculated")
```

To study convergence for `bmi` alone:
```{r}
plot(imp, c("bmi"))
```

Although the relation of `bmi` is preserved now in the imputations we get absurd imputations and on top of that we clearly see there are some problems with the convergence of `bmi`. The problem is that we have circularity in the imputations. We used passive imputation for `bmi` but `bmi` is also automatically used as predictor for `wgt` and `hgt`. This can be solved by adjusting the predictor matrix.

---

**7. Solve the problem of circularity (if you did not already do so) and plot once again the imputed values of bmi versus the calculated values.**

First, we remove `bmi` as a predictor for `hgt` and `wgt` to remove circularity.
```{r}
pred<-ini$pred
pred
pred[c("hgt", "wgt"), "bmi"] <- 0
pred
```
and we run the `mice` algorithm again with the new predictor matrix (we still 'borrow' the imputation methods object `meth` from before)
```{r}
imp <-mice(boys, meth=meth, pred=pred, print=FALSE)
```
Second, we recreate the plots from **Assignment 6**. We start with the plot to inspect the relations in the observed and imputed data
```{r}
xyplot(imp, bmi ~ I(wgt / (hgt / 100)^2), na.groups = miss,
       cex=c(1, 1), pch=c(1, 20),
       ylab="BMI (kg/m2) Imputed", xlab="BMI (kg/m2) Calculated")
```

and continue with the trace plot to study convergence
```{r}
plot(imp, c("bmi"))
```

All is well now!

---

**Conclusion**

We have seen that *post-processing* of imputations and *passive imputation* can be used to preserve typical features of the data while maintaining the relations between the variables in both the observed and imputed data.

It is important to ‘gain’ this control as a user. After all, we are imputing values and taking their uncertainty properly into account.

---

**For fun: what you shouldn’t do with passive imputation**

Never set all relations fixed. You will remain with the starting values and waste your computer’s energy (and your own).
```{r}
ini <- mice(boys, maxit = 0)
meth<- ini$meth
pred <- ini$pred
pred
meth["bmi"]<- "~ I(wgt/(hgt/100)^2)"
meth["wgt"]<- "~ I(bmi*(hgt/100)^2)"
meth["hgt"]<- "~ I(sqrt(wgt/bmi)*100)"
imp.path <- mice(boys, meth = meth, pred = pred, seed = 123)
plot(imp.path, c("hgt", "wgt", "bmi"))
```

We named the `mids`- object `imp.path`, because the nonconvergence is pathological in this example!

---

**- End of Vignette**

---
